Temporal ordering and registration of images in studies of developmental dynamics

### 2014-08-20 image_analysis_paper2/pnas_draft.tex
pnas paper edits

\begin{figure*}[t]
\includegraphics{fig3}
\caption{{\xfigtextfontit Drosophila} embryos during cellularization. {\xfigtextfontit (A)} Images of embryos collected during cellularization. Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation. {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
\customlabel{fig:data1}{3}
\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
\end{figure*}

\subsection{Ordering Images From a Movie}

As a first illustration, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
%
By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
%
\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
%
During the third hour of development, this arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
%
One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.

Throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (reviewed in \cite{rushlow2012temporal}).
%
The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the third hour of development \cite{lim2013kinetics}.
%
This bimodal pattern is then augmented by a peak at the dorsal side of the embryo.
%
Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see \fig~\ref{subfig:ordered_data1}).
%
The peak of nuclear Dl is consistent throughout the images and the intensity of the dpERK signal in the lateral regions of the embryo increases monotonically across the data set.
%

For this set of images, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
%
During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
%
The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 2--3 minutes \cite{figard2013plasma} (see \fig~\ref{fig:membrane_curve}).
%
The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
%
Thus, we have validated a dimensionality reduction approach to temporal ordering and registration on a data set where image orientation and ordering can be assessed independently.

@@

Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~\ref{fig:PCA_data1}).
%
However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure.
%
As an example, \fig~\ref{subfig:raw_data2} shows a set of $108$ images, which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
During this time, the arrangement of newly formed cells begins to deform into structures which will form the future organs \cite{leptin2005gastrulation}.
%
As before, each image is an optical cross-section of a vertically oriented embryo fixed at a different developmental time.
%
The nuclei (gray) are again labeled with a DNA marker.
%
However, during this developmental time period, the nuclei no longer remain uniformly arranged around the periphery of the embryo.
%
Instead of Dl, which is difficult to visualize at this point, embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
%
dpERK is again shown in red.

%
During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells \cite{leptin2005gastrulation}.
%
dpERK appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.

\begin{figure*}[t]
\includegraphics{fig4}
\caption{{\xfigtextfontit Drosophila} embryos during gastrulation. {\xfigtextfontit (A)} Images of gastrulating embryos. Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
\customlabel{fig:data2}{4}
\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
\end{figure*}

Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
%
We found that simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
%
\fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
and visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial;
and second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see \SI and  \cite{kemelmacher2011exploring}).  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

@@

We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots. 
%
The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
%
The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information.

@@

@@

%\bibliographystyle{pnas2}
%\bibliography{background_reading/references,../../references/references}

\begin{thebibliography}{10}

\bibitem{diggle2002analysis}
Diggle P, Heagerty P, Liang K~Y, Zeger S (2002) \textit{Analysis of
  longitudinal data} (Oxford University Press).

\bibitem{roelens2013live}
Roelens B, Chanes J~D~L~H, Dostatni N (2013) Live imaging of bicoid-dependent
  transcription in \textit{{D}rosophila} embryos. \textit{Curr Biol}
  23:1--5.

\bibitem{keller2013imaging}
Keller P~J (2013) Imaging morphogenesis: technological advances and biological
  insights. \textit{Science} 340(6137).

\bibitem{jaeger2004dynamic}
Jaeger J, Surkova S, Blagov M, Janssens H, Kosman D, \textit{et~al.} (2004)
  Dynamic control of positional information in the early \textit{{D}rosophila}
  embryo. \textit{Nature} 430(6997):368--371.

%\bibitem{peter2012predictive}
%Peter IS, Faure E, Davidson EH (2012) Predictive computation of genomic %logic processing functions in embryonic development. \textit{Proc Natl %Acad Sci USA} 109(41):16434--16442.

\bibitem{peter2011gene}
Peter IS, Davidson EH (2011) A gene regulatory network controlling the embryonic
specification of endoderm. \textit{Nature} 474(7353):635--639.

\bibitem{fowlkes2008quantitative}
Fowlkes C~C, Hendriks C~L~L, Ker{\"a}nen S~V, Weber G~H, R{\"u}bel O,
  \textit{et~al.} (2008) A quantitative spatiotemporal atlas of gene expression
  in the \textit{{D}rosophila} blastoderm. \textit{Cell} 133(2):364--374.

\bibitem{ng2012large}
Ng L~L, Sunkin S~M, Feng D, Lau C, Dang C, \textit{et~al.} (2012) Large-scale
  neuroinformatics for in situ hybridization data in the mouse brain.
  \textit{Int Rev Neurobiol} 104:159--182.

\bibitem{richardson2014emage}
Richardson L, Stevenson P, Venkataraman S, Yang Y, Burton N, \textit{et~al.}
  (2014) Emage: Electronic mouse atlas of gene expression, \textit{Mouse
  Molecular Embryology} (Springer), pp. 61--79.

\bibitem{castro2009automatic}
Castro C, Luengo-Oroz M, Desnoulez S, Duloquin L, Fernandez-de Manuel L,
  \textit{et~al.} (2009) An automatic quantification and registration strategy
  to create a gene expression atlas of zebrafish embryogenesis,
  \textit{Conf Proc IEEE Eng Med Biol Soc}, pp. 1469--1472.

\bibitem{zitova2003image}
Zitova B, Flusser J (2003) Image registration methods: a survey. \textit{Image Vis Comput} 21(11):977--1000.

\bibitem{rowley1998rotation}
Rowley H~A, Baluja S, Kanade T (1998) Rotation invariant neural network-based
  face detection, \textit{Computer Vision and Pattern Recognition, 1998.
  Proceedings. 1998 IEEE Computer Society Conference on} (IEEE), pp. 38--44.

\bibitem{hajnal2010medical}
Hajnal J~V, Hill D~L (2010) \textit{Medical image registration} (CRC press).

\bibitem{greenspan1994rotation}
Greenspan H, Belongie S, Goodman R, Perona P (1994) Rotation invariant texture
  recognition using a steerable pyramid, \textit{Pattern Recognition, 1994.
  Vol. 2-Conference B: Computer Vision \& Image Processing., Proceedings of the
  12th IAPR International. Conference on} (IEEE), volume~2, pp. 162--167.

\bibitem{zhao2003face}
Zhao W, Chellappa R, Phillips P~J, Rosenfeld A (2003) Face recognition: A
  literature survey. \textit{ACM Computing Surveys (CSUR)} 35(4):399--458.

\bibitem{singer2012vector}
Singer A, Wu H~T (2012) Vector diffusion maps and the connection laplacian.
  \textit{Commun Pure Appl Math} 65(8):1067--1144.

\bibitem{Belkin2003}
Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
  data representation. \textit{Neural Comput} 15:1373--1396.

\bibitem{coifman2005geometric}
Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
  Geometric diffusions as a tool for harmonic analysis and structure definition
  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.

\bibitem{coifman2006geometric}
Coifman R~R, Lafon S (2006) Geometric harmonics: a novel tool for multiscale
  out-of-sample extension of empirical functions. \textit{Appl Comput Harmon Anal} 21(1):31--52.

\bibitem{tenenbaum2000global}
Tenenbaum J~B, De~Silva V, Langford J~C (2000) A global geometric framework for
  nonlinear dimensionality reduction. \textit{Science} 290(5500):2319--2323.

\bibitem{roweis2000nonlinear}
Roweis S~T, Saul L~K (2000) Nonlinear dimensionality reduction by locally
  linear embedding. \textit{Science} 290(5500):2323--2326.

\bibitem{zhao2014rotationally}
Zhao Z, Singer A (2014) Rotationally invariant image representation for
viewing direction classification in cryo-EM. \textit{J Struct Biol} 186(1):153--166.

\bibitem{singer2011viewing}
Singer A, Zhao Z, Shkolnisky Y, Hadani R (2011) Viewing angle
classification of cryo-electron microscopy images using eigenvectors. \textit{SIAM J Imaging Sci} 4(2):723--759.

\bibitem{lafon2006data}
Lafon S, Keller Y, Coifman R~R (2006) 
Data fusion and multicue data matching by diffusion maps. \textit{IEEE Trans Pattern Anal Mach Intell}
28(11):1784--1797.

\bibitem{jaeger2012drosophila}
Jaeger J, Manu, Reinitz J (2012) \textit{{D}rosophila} blastoderm patterning.
  \textit{Curr Opin Genet Dev} 22(6):533 -- 541.

\bibitem{singer2011angular}
Singer A (2011) Angular synchronization by eigenvectors and semidefinite
  programming. \textit{Appl Comput Harmon Anal}
  30(1):20--36.

\bibitem{ahuja2007template}
Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
  relative equilibria in systems with continuous symmetry. \textit{Journal of
  Nonlinear Science} 17(2):109--143.

\bibitem{ian1998statistical}
Dryden I~L, Mardia K (1998) \textit{Statistical shape analysis}, Wiley series
  in probability and statistics: Probability and statistics (J. Wiley).

\bibitem{foe1983studies}
Foe V~E, Alberts B~M (1983) Studies of nuclear and cytoplasmic behaviour during
  the five mitotic cycles that precede gastrulation in \textit{{D}rosophila}
  embryogenesis. \textit{J Cell Sci} 61(1):31--70.

\bibitem{rushlow2012temporal}
Rushlow C~A, Shvartsman S~Y (2012) Temporal dynamics, spatial range, and
  transcriptional interpretation of the dorsal morphogen gradient.
  \textit{Curr Opin Genet Dev} 22(6):542--546.

\bibitem{lim2013kinetics}
Lim B, Samper N, Lu H, Rushlow C, Jim{\'e}nez G, \textit{et~al.} (2013)
  Kinetics of gene derepression by erk signaling. \textit{Proc Natl Acad Sci USA} 110(25):10330--10335.

\bibitem{figard2013plasma}
Figard L, Xu H, Garcia H~G, Golding I, Sokac A~M (2013) The plasma membrane
  flattens out to fuel cell-surface growth during \textit{{D}rosophila}
  cellularization. \textit{Dev Cell} 27(6):648--655.

\bibitem{shlens2005tutorial}
Shlens J (2005) A tutorial on principal component analysis. \textit{Systems
  Neurobiology Laboratory, University of California at San Diego} .

\bibitem{leptin2005gastrulation}
Leptin M (2005) Gastrulation movements: the logic and the nuts and bolts.
  \textit{Dev Cell} 8(3):305--320.

\bibitem{kemelmacher2011exploring}
Kemelmacher-Shlizerman I, Shechtman E, Garg R, Seit, S~M (2011)
Exploring photobios. \textit{ACM Transactions on Graphics (TOG)} 30(4):61.

\bibitem{snail}
Ip Y~T, Park R~E, Kosman D, Bier E, Levine M (1992) 
The dorsal gradient morphogen regulates stripes of rhomboid expression in the presumptive neuroectoderm of the \textit{Drosophila} embryo. \textit{Genes Dev} 6(9):1728--1739.

\bibitem{yuan2014automated}
Yuan L, Pan C, Ji S, McCutchan M, Zhou Z~H, \textit{et~al.} (2014) Automated
  annotation of developmental stages of \textit{{D}rosophila} embryos in images
  containing spatial patterns of expression. \textit{Bioinformatics}
  30(2):266--273.

\bibitem{surkova2008characterization}
Surkova S, Kosman D, Kozlov K, Myasnikova E, Samsonova A~A, \textit{et~al.}
  (2008) Characterization of the \textit{{D}rosophila} segment determination
  morphome. \textit{Dev Biol} 313(2):844--862.

\bibitem{anavy2014blind}
Anavy L, Levin M, Khair S, Nakanishi N, Fernandez-Valverde S~L, \textit{et~al.}
  (2014) Blind ordering of large-scale transcriptomic developmental
  timecourses. \textit{Development} 141(5):1161--1166.

\bibitem{trapnell2014dynamics}
Trapnell C, Cacchiarelli D, Grimsby J, Pokharel P, Li S, \textit{et~al.} (2014)
  The dynamics and regulators of cell fate decisions are revealed by
  pseudotemporal ordering of single cells. \textit{Nat Biotechnol} 32(4):381--386.

\bibitem{gupta2008extracting}
Gupta A, Bar-Joseph Z (2008) Extracting dynamics from static cancer expression
  data. \textit{IEEE/ACM Trans Comput Biol Bioinform} 5(2):172--182.

%\bibitem{singer2011three}
%Singer A, Shkolnisky Y (2011) Three-dimensional structure determination from
%  common lines in cryo-em by eigenvectors and semidefinite programming.
%  \textit{SIAM J Imaging Sci} 4(2):543--572.

%\bibitem{silverman2005functional}
%Ramsay J~O, Silverman B~W (2005) \textit{Functional Data Analysis}, Springer
%  Series in Statistics (Springer).

\bibitem{goemans1995improved}
Goemans M~X, Williamson D~P (1995) Improved approximation algorithms for maximum cut and satisfiability problems using semidefinite programming.
\textit{Journal of the ACM} 42(6):1115--1145.

\bibitem{bandeira2013cheeger}
Bandeira A~S, Singer A, Spielman D~A (2013) A Cheeger inequality for the graph connection Laplacian. \textit{SIAM Journal on Matrix Analysis and Applications} 34(4):1611--1630.

\bibitem{mallat2012group}
Mallat, S (2012) Group invariant scattering. 
\textit{Communications on Pure and Applied Mathematics}
65(10):1331--1398.

\bibitem{krzic2012multiview}
Krzic U, Gunther S, Saunders T~E, Streichan S~J, Hufnagel L (2012) Multiview
  light-sheet microscope for rapid in toto imaging. \textit{Nat Methods}
  9(7):730--733.

\bibitem{ichikawa2014live}
Ichikawa T, Nakazato K, Keller P~J, Kajiura-Kobayashi H, Stelzer E~H,
  \textit{et~al.} (2014) Live imaging and quantitative analysis of gastrulation
  in mouse embryos using light-sheet microscopy and 3d tracking tools.
  \textit{Nat Protoc} 9(3):575--585.

\bibitem{rubel2010coupling}
R{\"u}bel O, Ahern S, Bethel E, Biggin M~D, Childs H, \textit{et~al.} (2010)
  Coupling visualization and data analysis for knowledge discovery from
  multi-dimensional scientific data. \textit{Procedia Comput Sci}
  1(1):1757--1764.

\end{thebibliography}

@@

%\begin{figure}[t]
%\includegraphics{fig1}
%\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%\label{fig:fish}
%\customlabel{fig:fish}{1}
%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
%\end{figure}
%
%\begin{figure}[t]
%\includegraphics{fig2}
%\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%\label{fig:schematics}
%\customlabel{fig:schematics}{2}
%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
%\end{figure}
%
%\begin{figure*}[t]
%\includegraphics{fig3}
%\caption{{\xfigtextfontit Drosophila} embryos during cellularization. {\xfigtextfontit (A)} Images of embryos collected during cellularization. Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation. {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
%\customlabel{fig:data1}{3}
%\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
%\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
%\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
%\end{figure*}
%
%\begin{figure*}[t]
%\includegraphics{fig4}
%\caption{{\xfigtextfontit Drosophila} embryos during gastrulation. {\xfigtextfontit (A)} Images of gastrulating embryos. Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
%\customlabel{fig:data2}{4}
%\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
%\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
%\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
%\end{figure*}
%
%\begin{figure}[h]
%\includegraphics{fig5}
%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.  {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates. The directions of the developmental trajectories are indicated by arrows. (Inset) The projected data, now colored by genotype. The two branches separate mutant and wild type samples. {\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The two dpERK peaks which arise during development are indicated by arrows. {\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The single dpERK peak is indicated by an arrow. }
%\customlabel{fig:data3}{5}

>>>>

\begin{figure*}[t]
\includegraphics{fig3}
\caption{{\xfigtextfontit Drosophila} embryos during cellularization. {\xfigtextfontit (A)} Images of embryos collected during cellularization. Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation. {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
\customlabel{fig:movie}{3}
\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}{\it A}}
\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{\it B}}
\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{\it C}}
\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{\it D}}
\customlabel{subfig:movie_error_table}{\ref{fig:movie}{\it E}}
\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{\it F}}
\end{figure*}

\subsection{Method validation using a live movie}

As a first illustration, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by live imaging of a single {\it Drosophila} embryo during the thirty minute time interval spanning late cellularization through gastrulation.
%
During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells and eventually forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side.

The movie shows an optical cross-section of a vertically oriented developing embryo, where the nuclei are labeled by DAPI, a DNA stain.
%
The movie contains $42$ frames, taken at $30$~sec time intervals;
\fig~\ref{subfig:movie_unregistered_unordered} shows selected frames from this movie. 
%
The gray nuclear signal has been blurred to obtain a smoother images.
%
Each frame has been arbitrarily rotated, and the order of the frames has been scrambled.

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{subfig:movie_registered_ordered} shows the same selected frames, now registered and ordered.
%
The real time for each frame is indicated.
%
\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  shows the correlations between the recovered and true angles and ranks for each of the frames.
%
Both the angles and the ranks are recovered to a high degree of accuracy.

To assess the robustness of our proposed methodology, we repeated this procedure with four additional movies.
%
The results are shown in \fig~\ref{subfig:movie_error_table}. 
%
The errors in the recovered angles are all very low, and the rank correlation coefficients are consistently close to 1, indicating that our methodology can reproducably order data of this type.

To assess how the quality of our results depends on the number of data points, we used bootstrap sampling. 
%
\fig~\ref{subfig:movie_bootstrap} shows the average rank correlation coefficient as a function of the number of images in the data set.
%
The data sets were sampled (with replacement) from the original $40$-frame movie.
%
There is a steady increase in the accuracy of the recovered orderings with an increasing number of images.
%
The recovered angles had errors of less than $10^{\circ}$, and the magnitude of the errors did not vary significantly with the size of the data set.

@@

We will now demonstrate the utility of our methodogies on a data set where the correct ordering is not known {\em a priori}.
%
\fig~\ref{subfig:fixed_images_unregistered_unordered} shows a set of $126$ images, which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
Each image is an optical cross-section of a {\em different} vertically oriented embryo fixed at a different (and unknown) developmental time.
%
The nuclei (gray) are again labeled with a DNA marker.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.

\begin{figure*}[t]
\includegraphics{fig4}
\caption{{\xfigtextfontit Drosophila} embryos during gastrulation. {\xfigtextfontit (A)} Images of gastrulating embryos. Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
\customlabel{fig:data2}{4}
\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
\end{figure*}

FIX!!

\fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
and visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial;
and second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see \SI and  \cite{kemelmacher2011exploring}).  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

%
dpERK appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.

@@

\begin{itemize}
\item High--throughput imaging is becoming increasingly common
\item We are/wil reach the point where imaging data is being produced more rapidly than it can be analyzed by hand (similar to microarray data)
\item We need a way to automate the organization of such data
\item Our algorithms register, which is important for many imaging applications
\item In general our algorithms find good coordinates/descriptors for the data
\item We focus on uncovering a single coordinate that can be used to order data
\item in general, can uncover many coordinates; use coordinates to do other data processing tasks (clustering, model fitting, etc.)

\end{itemize}
We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots. 
%
The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
%
The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information.

We are approaching the time when the rate of imaging data collection will surpass the rate of manual image analysis. 
%
We therefore require automated methodologies to organize such large data sets; these methodogies should be sufficiently general to address a wide variety of applications. 
%
Vector diffusion maps allows us to automatically register images, which is an essential task for many applications.
%
Simultaneously the algorithm provides us with a parameterization or coordinates for each image.
%
In the examples presented here, we have focued on recovering a single coordinate, and using this coordinate to order the images, effectively ordering them in time.
%
In general, we can recover an arbitrariy number of coordinates, and use this parametersization to do any of the typical data analysis tasks, such as outlier detection, clustering, and model fitting.

@@

  \begin{itemize}
  \item Discuss parameter robustness/sensitity/tuning
  \item timing/scaling
  \item mention code
  \item SNR? Add Gaussian Noise?
  \end{itemize}

@@

\bibliographystyle{pnas2}
\bibliography{../image_analysis_paper/background_reading/references,../../references/references}

@@
